Score-based tests of measurement invariance: Use in practice

### 2014-01-30 frontiers/frontiers.Rnw
Edited half of the frontiers paper and comments.
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/semtools@311 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

Many of the papers in this special issue focus on the topic of approximate
measurement invariance: we know that strict hypotheses of measurement invariance
do not hold exactly across different groups, and this should be reflected in
corresponding tests of measurement invariance.  Under a Bayesian
approach, we may utilize the idea of approximate invariance
\cite<e.g.,>{muthen13}, whereby across-group equality constraints on
parameters are replaced with informative prior distributions.  In this paper, we
describe an alternative approach: the development of test statistics that 
are especially sensitive to violations that are monotonic w.r.t. the
variable of interest (and 
less sensitive to violations that are non monotonic w.r.t. the
variable of interest).  Because monotonic violations are more likely to be
interpretable and interesting to the researcher, we can gain more
power to detect these violations by de-emphasizing other types of violations.
The resulting test statistics are specifically applicable to
situations where one  
wishes to test for measurement invariance with respect to an ordinal variable, 
and they are special cases of a family of tests that may be used to study 
measurement invariance w.r.t.\ continuous, categorical, and ordinal variables
\cite{mer93,DolMaa98,lubmut05}\readme{The new references do not work
  here (because they are unrelated to tests based on stochastic
  processes).  I will work on these references.}.

The family of tests described in this paper have recently been applied
to the study of measurement invariance \cite{MerZei13,MerFanZei}, though their
practical application has been limited to a small set of simulations and
datasets.  In this paper, we provide a detailed illustration of the tests' use
and performance under
scenarios likely to be encountered in practice. 
%In order to fix the psychometric scale, we need to specify the violating parameter.  
%Moreover, in practice model misspecification is an inevitable issue to a certain extent 
%and may interfere with parameter invariance specificity. A family of tests derived from 
%stochastic process has been recently proposed to identify specific parameters impacted 
%by measurement invariance violation. However, these tests have only been used to a limited 
%type of parameter violation. It is this paper's intent to provide a detailed examination 
%of the tests' performance under practical research scenarios. 
In the following sections, we first briefly review the  theoretical framework of
the proposed tests and provide a short tutorial illustrating the use of the
tests in \proglang{R} \cite{R11}. Next, we study the tests' performance in simulations that mimic
practical research scenarios.  Finally, we provide some further discussion on
the tests' use in practice.

@@

To verbally describe the above equation, each
individual has $k$ scores describing the extent to which the fitted
model describes that particular individual.  We can roughly interpret
scores near zero as providing a ``good'' description of an individual,
with scores far from zero providing a ``bad'' description of an
individual.  This is only a rough interpretation as, even when measurement
invariance holds, some individuals' scores will be further from zero
than others.  Each of the $k$ scores represents one model parameter,
which, as further described below, allows us to test subsets of model
parameters for invariance.

@@

To formalize these ideas, we define a
suitably scaled cumulative sum of the ordered scores.  This may be written as
\begin{equation} 
    \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}
where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
nt \rfloor$ is 
the integer part of $nt$ (i.e., a floor 
operator), and $x_{(i)}$ reflects the individual with the $i$-th
smallest value of the auxiliary variable $V$.  While the above
equation is written in general form, we can
restrict the value of $t$ in finite samples to the set $\{0,
1/n, 2/n, 3/n, \ldots, n/n\}$.
We focus on how the
cumulative sum fluctuates as more individuals' scores are added to it, e.g.,
starting with the youngest and ending with the oldest individual
if age is the auxiliary variable of interest.
The summation is premultiplied by an estimate of the inverse square
root of the information matrix, which serves to decorrelate the
fluctuation processes associated with individual model parameters
while preserving the behavior of individual parameters' fluctuations..

@@

To conduct the analysis, we first load the data and keep only complete cases for simplicity 
(though the tests can be applied to incomplete data).
%
<<data>>=
data("YouthGratitude", package = "psychotools")
compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
yg <- YouthGratitude[compcases, ]
@
%
Next, we fit two models in \pkg{lavaan}: a one-factor model 
where loadings are restricted to be equal across age groups, 
and a one-factor model where loadings are free across age groups.
This allows us to test a hypothesis of partial measurement invariance
that was of interest to the original researchers (though, for ordinal
variables, all types of
measurement invariance can be examined via the tests described
previously).
By default, the code below sets the scale by fixing the first loading
to 1.
%
<<models>>=
restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
  data = yg, group = "agegroup", meanstructure = TRUE,
  group.equal = "loadings")

@@

To examine these issues, we generated data from a
two-factor model with three indicators each (see Figure~\ref{fig:path}).
The measurement invariance
violation occurred in one of four places: the factor loading associated with Scale 1 
 ($\lambda_{11}$), the unique variance  ($\psi_{11}$)
 or intercept associated with Scale 1 ($\mu_{11}$)
, the factor covariance ($\phi_{12}$).  Note that the
 latter violation is not necessarily a measurement invariance
 violation, but it is still a parameter instability that can occur in
 this type of model.
 We then tested for measurement invariance (parameter instability) in
 seven subsets of parameters: each of the four
individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 
all six unique variances and all six intercepts. % $\epsilon_{11}--\epsilon_{66}$.

>>>>

Many of the papers in this special issue focus on the topic of approximate
measurement invariance: we know that strict hypotheses of measurement invariance
do not hold exactly across different groups, and this should be reflected in
corresponding tests of measurement invariance.  Under a Bayesian
approach, we may utilize the idea of approximate invariance
\cite<e.g.,>{muthen13}, whereby across-group equality constraints on
parameters are replaced with informative prior distributions.  In this paper, we
describe an alternative approach: the development of test statistics that 
are especially sensitive to violations that are monotonic w.r.t. the
variable of interest (and 
less sensitive to violations that are non monotonic w.r.t. the
variable of interest).  Because monotonic violations are more likely to be
interpretable and interesting to the researcher, we can gain more
power to detect these violations by de-emphasizing other types of violations.
The resulting test statistics are specifically applicable to
situations where one  
wishes to test for measurement invariance with respect to an ordinal variable, 
and they are special cases of a family of tests that may be used to study 
measurement invariance w.r.t.\ continuous, categorical, and ordinal
variables.  Other methods for studying measurement
invariance w.r.t.\ continuous variables include moderated factor
models \cite{bauhus09,moldol10,pur02} and factor mixture models
\cite{dolmaa98,lubmut05}.  These methods require a model of greater
complexity to be estimated, while the tests described here work solely
on the output of a traditional factor model.  See \citeA{MerZei13} for
further comparison of these methods.

The family of tests described in this paper have recently been applied
to the study of measurement invariance \cite{MerZei13,MerFanZei}, though their
practical application has been limited to a small set of simulations and
datasets.  In this paper, we provide a detailed illustration of the tests' use
and performance under
scenarios likely to be encountered in practice.  While the previous
papers have described and studied the tests under ideal conditions, we
focus here on two topics of interest to the applied researcher:
software considerations for carrying out the tests, and test
performance under model misspecification.
%In order to fix the psychometric scale, we need to specify the violating parameter.  
%Moreover, in practice model misspecification is an inevitable issue to a certain extent 
%and may interfere with parameter invariance specificity. A family of tests derived from 
%stochastic process has been recently proposed to identify specific parameters impacted 
%by measurement invariance violation. However, these tests have only been used to a limited 
%type of parameter violation. It is this paper's intent to provide a detailed examination 
%of the tests' performance under practical research scenarios. 
In the following sections, we first briefly review the  theoretical framework of
the proposed tests and provide a short tutorial illustrating the use of the
tests in \proglang{R} \cite{R11}. Next, we study the tests' performance in simulations that mimic
practical research scenarios.  Finally, we provide some further discussion on
the tests' use in practice.

@@

To verbally describe the above equation, each
individual has $k$ scores describing the extent to which the fitted
model describes that particular individual.  We can roughly interpret
scores near zero as providing a ``good'' description of an individual,
with scores far from zero providing a ``bad'' description of an
individual.  This is only a rough interpretation as, even when measurement
invariance holds, some individuals' scores will be further from zero
than others.  However, under measurement noninvariance, the scores
will differ for different subgroups of individuals (say, scores in
subgroup A tend to be negative and scores in subgroup B tend to be
positive).  Each of the $k$ scores represents one model parameter,
which, as further described below, allows us to test subsets of model
parameters for invariance.

@@

To formalize these ideas, we define a
suitably scaled cumulative sum of the ordered scores.  This may be written as
\begin{equation} 
    \label{eq:cumscore}
  {\bm B}(t; \hat {\bm \theta}) ~=~ \hat {\bm I}^{-1/2} n^{-1/2}
    \sum_{i = 1}^{\lfloor n \cdot t \rfloor} {\bm s}(\hat {\bm \theta}; x_{(i)})
  \qquad (0 \le t \le 1)
\end{equation}
where $\hat{\bm I}$ is an estimate of the information matrix, $\lfloor
nt \rfloor$ is 
the integer part of $nt$ (i.e., a floor 
operator), and $x_{(i)}$ reflects the individual with the $i$-th
smallest value of the auxiliary variable $V$.  While the above
equation is written in general form, we can
restrict the value of $t$ in finite samples to the set $\{0,
1/n, 2/n, 3/n, \ldots, n/n\}$.
We focus on how the
cumulative sum fluctuates as more individuals' scores are added to it, e.g.,
starting with the youngest and ending with the oldest individual
if age is the auxiliary variable of interest.
The summation is premultiplied by an estimate of the inverse square
root of the information matrix, which serves to decorrelate the
fluctuation processes associated with individual model parameters
while preserving the behavior of individual parameters' fluctuations.

@@

To conduct the analysis, we first load the data and keep only complete cases for simplicity 
(though the tests can be applied to incomplete data).
%
<<data>>=
data("YouthGratitude", package = "psychotools")
compcases <- apply(YouthGratitude[, 4:28], 1, function(x) all(x %in% 1:9))
yg <- YouthGratitude[compcases, ]
@
%
Next, we fit two models in \pkg{lavaan}: a one-factor model 
where loadings are restricted to be equal across age groups, 
and a one-factor model where loadings are free across age groups.
This allows us to test a hypothesis of weak measurement invariance
that was of interest to the original researchers (though, for ordinal
variables, all types of
measurement invariance can be examined via the tests described
previously).
By default, the code below sets the scale by fixing the first loading
to 1.
%
<<models>>=
restr <- cfa("f1 =~ gq6_1 + gq6_2 + gq6_3 + gq6_4 + gq6_5",
  data = yg, group = "agegroup", meanstructure = TRUE,
  group.equal = "loadings")

@@

To examine these issues, we generated data from a
two-factor model with three indicators each (see Figure~\ref{fig:path}).
The measurement invariance
violation occurred in one of four places: the factor loading associated with Scale 1 
 ($\lambda_{11}$), the unique variance  ($\psi_{11}$)
 or intercept associated with Scale 1 ($\mu_{11}$)
, the factor covariance ($\phi_{12}$).  Note that the
 latter violation is not necessarily a measurement invariance
 \cite<e.g.,>{mer93}, but it is still a parameter instability that can occur in
 this type of model.
 We then tested for measurement invariance (parameter instability) in
 seven subsets of parameters: each of the four
individual parameters noted above, all six factor loadings, % $\lambda_{11}--\lambda_{62}$, and 
all six unique variances and all six intercepts. % $\epsilon_{11}--\epsilon_{66}$.
